Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-07-22 text/manuscript-full-KS.qmd
last to dos

[^10]: Quantification often reduces the decision-making authority of local actors [@RN5007]. The Market Segment Model sought to replicate the aggregate knowledge of local admissions officers. Once this knowledge was quantified and commodified onto a CD-ROM, the local expertise of admissions officers becomes less valuable. The EPS product increases the ability of a college admissions leader -- working with College Board staff or an EM consultant -- to plan recruiting efforts centrally. In background conversations, EM professionals told us that EPS software enabled colleges to plan travel without relying on admissions officers having strong local knowledge of their territories. However, EPS software often recommended visiting the same sets of affluent, high-achieving high schools that were receiving visits from other colleges. We were told that savvier, quantitatively-adept admissions offices used EPS to visit schools that EPS recommended ignoring because there would be less competition for the good students who attended these high schools. Finally, Geomarkets were created to mirror the territories of admissions officers [@RN4982]. In turn, as Geomarkets became more salient, admissions offices often structured their territories around Geomarkets. On background, admissions officers and enrollment consultants told us it remains common parlance to hear an admissions officer say something like, "I recruit 'PA 2'," which refers to the "Chester County, PA Geomarket." In states a college recruits heavily from, admissions officers are often assigned specific Geomarkets as their territory. This can be seen in the Chicago-land recruiting territories of University of Chicago, Lake Forest College, and Saint Mary's College. <!--\[ADD TO APPENDIX CRYSTAL\]-->

@@

The primary data source for our second research question is student lists purchased from College Board for which we received both (a) the order summary data and (b) the de-identified prospect-level student list data. This sample includes a total of 414 orders associated with 2,549,085 prospects made by 14 public universities located in California, Illinois, Texas, and Arizona. Although these lists were purchased by individual universities, the set of prospects included in each list is a function of the search criteria specified for that student list purchase. The resulting student lists could hypothetically extend to any university placing student lists orders based on the same filter criteria. In other words, any observed patterns in the findings below would remain constant for any university selecting the same filters.[^18] Therefore, we utilize these data not to analyze the behavior of individual universities, but to identify which prospective students are included when a particular set of search criteria are selected.

@@

::: {.landscape}

```{r fig-chicago-recruiting, echo = FALSE, fig.align = 'center', fig.cap = "The College Board Enrollment Planning Service Chicago Geomarkets (left) and University of Chicago Recruiting Territories (right)", out.width = "100%"}
knitr::include_graphics('./../assets/images/chicago_recruiting.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

:::

@@

::: {.landscape}

```{r fig-geomarket-methods, echo=FALSE,out.width="100%", fig.cap="caption",fig.align='center',fig.cap = "Chicago, College Board and Recreated Geomarkets"}
knitr::include_graphics('./../assets/images/eps/chicago_geos.png')

@@

```{r fig-chicago-maps, echo = FALSE, fig.align = 'center', fig.cap = "Chicago Geomarkets, Total Population 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/chicago_maps.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

:::

@@

::: {.landscape}

```{r fig-dallas-maps, echo = FALSE, fig.align = 'center', fig.cap = "Dallas-Forth Worth Geomarkets, Total Population 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/dallas_maps.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

:::

@@

::: {.landscape}

```{r fig-la-maps, echo = FALSE, fig.align = 'center', fig.cap = "Los Angeles Geomarkets, Total Population 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/la-maps.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

:::

>>>>

[^10]: Quantification often reduces the decision-making authority of local actors [@RN5007]. The Market Segment Model sought to replicate the aggregate knowledge of local admissions officers. Once this knowledge was quantified and commodified onto a CD-ROM, the local expertise of admissions officers becomes less valuable. The EPS product increases the ability of a college admissions leader -- working with College Board staff or an EM consultant -- to plan recruiting efforts centrally. In background conversations, EM professionals told us that EPS software enabled colleges to plan travel without relying on admissions officers having strong local knowledge of their territories. However, EPS software often recommended visiting the same sets of affluent, high-achieving high schools that were receiving visits from other colleges. We were told that savvier, quantitatively-adept admissions offices used EPS to visit schools that EPS recommended ignoring because there would be less competition for the good students who attended these high schools. Finally, Geomarkets were created to mirror the territories of admissions officers [@RN4982]. In turn, as Geomarkets became more salient, admissions offices often structured their territories around Geomarkets. On background, admissions officers and enrollment consultants told us it remains common parlance to hear an admissions officer say something like, "I recruit 'PA 2'," which refers to the "Chester County, PA Geomarket." In states a college recruits heavily from, admissions officers are often assigned specific Geomarkets as their territory. This can be seen in the Chicago-land recruiting territories of University of Chicago in @fig-chicago-recruiting. <!--\[ADD TO APPENDIX CRYSTAL\]-->

@@

The primary data source for our second research question is student lists purchased from College Board. Originally collected from a larger project sample (Author, BLINDED), the data for this study's sample includes a total of 414 orders associated with 2,549,085 prospects made by 11 public universities located in California, Illinois, Texas, and Arizona. Although these lists were purchased by individual universities, the set of prospects included in each list is a function of the search criteria specified for that student list purchase. The resulting student lists could hypothetically extend to any university placing student lists orders based on the same filter criteria. In other words, any observed patterns in the findings below would remain constant for any university selecting the same filters.[^18] Therefore, we utilize these data not to analyze the behavior of individual universities, but to identify which prospective students are included when a particular set of search criteria are selected.

@@

::: landscape
```{r fig-chicago-recruiting, echo = FALSE, fig.align = 'center', fig.cap = "The College Board Enrollment Planning Service Chicago Geomarkets (left) and University of Chicago Recruiting Territories (right)", out.width = "100%"}
knitr::include_graphics('./../assets/images/chicago_recruiting.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```
:::

@@

::: landscape
```{r fig-geomarket-methods, echo=FALSE,out.width="100%", fig.cap="caption",fig.align='center',fig.cap = "Chicago, College Board and Recreated Geomarkets"}
knitr::include_graphics('./../assets/images/eps/chicago_geos.png')

@@

```{r fig-chicago-maps, echo = FALSE, fig.align = 'center', fig.cap = "Chicago Geomarkets, Total Population 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/chicago_maps.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```
:::

@@

::: landscape
```{r fig-dallas-maps, echo = FALSE, fig.align = 'center', fig.cap = "Dallas-Forth Worth Geomarkets, Total Population 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/dallas_maps.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```
:::

@@

::: landscape
```{r fig-la-maps, echo = FALSE, fig.align = 'center', fig.cap = "Los Angeles Geomarkets, Total Population 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/la-maps.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```
:::
